Enhance Road Detection Data Processing of LiDAR Point Clouds to Specifically Identify Unmarked Gravel Rural Roads

### 2023-05-18 Huston_Paper.tex
05182023

	{Gravel roads comprise 34.8\% \cite{road_stats_2} of all road surfaces in the United States. Some nations have predominately unpaved road networks, such as India, where 70.7\% of all roads by mile are categorized as unpaved \cite{malik_lal_2019}. Detection of gravel roads may allow for expansion of current civilian autonomous vehicle capabilities as well as introduce opportunities for industries that operate in rural areas. Global Positioning Service (GPS) and high resolution maps cannot be relied upon for navigation of rural roads alone as GPS is not accurate enough or may not have a reliable signal for precision driving \cite{noauthor_gpsgov_nodate}, and gravel roads are not marked or mapped. Systems that rely on cameras may fail when analyzing gravel roads, as these lack visual cues such as painted lane markings \cite{crisman_scarf_1993} and may in some circumstances closely resemble surrounding terrain. LiDAR point cloud processing systems depend upon distinct geometric features, most commonly curbs \cite{yadav_extraction_2017,liu_new_2013,qiu_fast_2016,fernandes_road_2014,seker_experiments_nodate,yang_semi-automated_2013,miyazaki_line-based_2014,hervieu_road_2013,smadja_road_nodate}, however as curbs are not installed on rural gravel roads \cite{skorseth_gravel_nodate} this method breaks down. Normal Distribution Transform (NDT) scan matching analytically compares two point cloud data sets to track current position, however this model relies on distinct geometric features, and suffers when distinguishing between terrain types making it insufficient for determining road boundaries \cite{biber_normal_2003}. Using 2D projection of planes unto a 3D point cloud of a road surface is an alternative approach, but likewise requires curbs or painted lines for road boundary definition \cite{fernandes_road_2014, borkar_robust_2009-1, guo_lane_2015}. Instead, using LiDAR may overcome difficulties in achieving detection of gravel roads by using terrain classification.}

	{Detection of unmarked road surfaces may utilize roughness properties without distinct markings or topographical features. Surface roughness is a measurable property that may be used to characterize a gravel road, as they are typically consistent with gravel type used \cite{skorseth_gravel_nodate} which may be exploited in a search and compare function, and is distinct from chipseal, grass, and other ground types \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain}. Surface roughness properties may be derived from processing LiDAR spatial and remission data from an aerial and surface level perspective \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain, pollyea_experimental_2012,rychkov_computational_2012,lague_accurate_2013,brubaker_use_2013,turner_estimation_2014,campbell_lidar-based_2017,shepard_roughness_2001,tegowski_statistical_2016,sock_probabilistic_2016,milenkovic_roughness_2018,yadav_extraction_2017, yadav_rural_2018}. Implementing a terrain classification model using Random Decision Forests (RDF) using scanning LiDAR data to exploit surface roughness was studied for road surface detection in this study. Evaluation of the detection model will rely principally upon accuracy and processing time, as a moving vehicle must obtain reliable information of the road surface in a timely manner in order to inform trajectory updates, or else risk suffering an accident.}

@@

	{Current LiDAR based road surface detection models may be categorized into two methodologies. Firstly is the analysis of road surface properties, which considers aspects such as topology and surface roughness. Detection based on road surface smoothness was studied by segmenting the point cloud into candidate road surface regions and searching for elevation jumps \cite{liu_new_2013}. Secondly is the detection of roadside curbs, which relies upon the height difference between the road surface and the curb for road edge detection. LiDAR point density at a curb's upper and lower edges can be used to indicate road edges \cite{ibrahim_curb-based_2012}. Fernandes et al. propose a road surface detection method using LiDAR \cite{fernandes_road_2014}, by projecting a 2D reference plane unto the 3D LiDAR data. In this study, road surfaces are assumed to be flat regions between two elevated regions such as curbs. Cameras have been used to supplement LiDAR point cloud data, which can implement 3D colored elevation maps to match pre-stored geometry data in a modified Iterative Closest Point (ICP) algorithm \cite{manz_detection_2011}. Supervised Classification Applied to Road Following (SCARF) is an algorithm that detects road surfaces based on color differences between the road and the surrounding terrain \cite{crisman_scarf_1993}. It was found that difficulties arise when detecting road surfaces in less colored environments \cite{crisman_scarf_1993,manz_detection_2011}. Edmond et al. propose a method to detect terrain type through implementation of vehicle vibration, however this method is limited to what the vehicle is currently driving over and thus cannot detect intercepting gravel roads \cite{dupont_online_2008}.}

@@

			{Followed methodology to accomplish this work is outlined in Fig. \ref{fig:flowz_2}. Training data was gathered by the Ohio University Autonomous Van van (Fig. \ref{fig:Experimental_Apperatus}). MATLAB was used to extract and parse LiDAR and GPS data. Scanning LiDAR data was aggregated into a large point cloud by using GNSS and IMU data to determine each LiDAR scan's position and orientation in a dead-reckoning process. Aggregated point cloud data was manually examined and compared to satellite and camera images to determine and project road boundaries. Training data was extracted from a single arc per 360 degree scanning LiDAR channel if the arc was coincident to a manually projected 2D training data area. Random Decision Forests were trained and verified using MATLAB's Classifier App. Scanning LiDAR data was consecutively classified and compared to manually defined truth areas yielded the true-positive accuracy for gravel, asphalt, and non-road surfaces. Classified point cloud data was examined without reference to truth areas in order to manually project possible road surfaces. Manually projected road surface areas were compared to truth areas to determine the accuracy of asphalt road and intercepting gravel driveway detection.}

@@

			{Required physical data was gathered by the Ohio University Autonomous Van (Fig. \ref{fig:Experimental_Apperatus}). LiDAR point cloud data was gathered with the vehicle's roof-mounted Velodyne VLP-32C. Velodnyne's VLP-32C has 32 channels producing 300,000 points per second with a vertical field of view from -45$^{\circ}$ to $+$15$^{\circ}$, providing a high output data stream \cite{vlp_32c}. GPS data was recorded with a Novatel PwrPak 7D-E2, a Global Navigation Satellite System (GNSS) \& Inertial Navigation System (INS), and two GNSS-502 antennas by NavtechGPS. Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using ROS. Scanning LiDAR data of a gravel surface was gathered from a gravel parking lot that intersects Blackburn Road near Athens Ohio. Scanning LiDAR data of an asphalt surface was gathered from Blackburn Road near the intersecting gravel lot. Scanning LiDAR data of grassy surfaces was gathered from the side of the gravel lot. Although this work does not evaluate the detection accuracy of grass, differentiation between a road and non-road surface is necessary for evaluating road-surface detection, therefore grass is labeled as "unknown" in this work. Raw data was gathered using the Robotic Operating System (ROS) to package incoming scanning LiDAR, GNSS, IMU, and camera sensor data as rosbags. Rosbags were unpacked using MATLAB tools to separate data streams. Point cloud, GPS, and IMU topics were defined, and point cloud Cartesian coordinates with corresponding intensities, longitude, latitude, altitude, roll, pitch, and yaw messages were read from the appropriate topics and imported}

@@

			{Scanning LiDAR training data requires extraction from manually defined gravel, asphalt, and "unknown" terrain areas on aggregated point clouds as manual extraction from consecutive scans is extremely time consuming. Aggregating LiDAR data into a single point cloud was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road. Scanning LiDAR timestamps were matched to the closest GNSS and IMU timestamp data. Closest matching GNSS and IMU data informed the location and orientation of the LiDAR scan. Point cloud translation and rotation was accomplished using transformation matrices derived from GPS and IMU data. Rotation matrices were created using extracted roll, pitch, and yaw data from the IMU. LiDAR origin was found using GPS longitude, latitude, and altitude data. Physical distances between the GPS, IMU, and LiDAR sensors was rectified by obtaining reference frames provided by AutonomouStuff. GPS coordinates were offset by the current orientation and converted the ground truth to the LiDAR frame. GPS, IMU, and LiDAR reference frames and rotational updates were combined into trajectory vectors. Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Fig. \ref{fig:road_areas_annotated_22}). While not as robust as more sophisticated methods of point cloud aggregation such as NDT or ICP scan matching, over shorter distances using this method proved to be adequate for this work.}

			{Manually defined areas may then be overlayed unto the point cloud (Fig. \ref{fig:road_areas_annotated_22}). Six degree arcs from the first three channels of scanning LiDAR data that lay directly in front of the vehicle were extracted and exported to a database if coincident with the manually defined areas. Training data was extracted from a single arc from each 360 degree LiDAR scan if the arc was coincident to a manually projected 2D area. Raw spatial and intensity values from the coincident arc were saved to a raw training database. Separate training databases were maintained for each individual channel. Features that describe the raw spatial and intensity data were extracted. Thirty percent of the training data base was randomly selected and set aside as a validation data base. Training data was visually examined for clustering to verify that a Random Decision Forest would be able to be trained (Fig. \ref{fig:training_data_cluster}). It was found that each class generally clustered to a certain range with some overlap, which indicates that a Random Decision Forest may be trained.}

			\begin{figure}[H]
				\centering
				\includegraphics[width=0.75\linewidth]{figures/vlcsnap-2023-04-20-08h48m17s447}
				\caption[Blackburn Road Camera View]{Camera view of Blackburn Road - an unmarked asphalt road with three intercepting gravel driveways from which asphalt data was gathered.}
				\label{fig:Blackburn_Road_View}
			\end{figure}

			\begin{figure}[H]
				\centering
				\includegraphics[width=0.75\linewidth]{figures/gravel_training_lot}
				\caption[Gravel Training Lot]{Camera view of the gravel lot from which gravel training data was gathered.}
				\label{fig:gravel_training_lot}
			\end{figure}

			\begin{figure}[H]
				\centering
				\includegraphics[width=0.75\linewidth]{figures/road_areas_annotated_22}
				\caption[Compiled and Manually Classified Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map. Projected manually defined areas define terrain surface type. Area $3$ is the gravel parking lot entrance from which training data was extracted.}
				\label{fig:road_areas_annotated_22}
			\end{figure}

			\begin{figure}[H]
			\centering
			\includegraphics[width=0.75\linewidth]{figures/training_data_cluster}
			\caption[Example Clustering]{Example of training data comparison between two features.}
			\label{fig:training_data_cluster}
			\end{figure}

@@

			{Feature extraction was completed by performing mathematical functions on the gathered spatial training data. Standard Deviation ($\sigma$), Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), where $G$ is an $1*n$ array comprising of the differences between consecutive points in the spatial data array. Spatial features were non-dimensionalized as necessary by dividing by the average range of the arc to LiDAR point of origin to the appropriate power. Remission features include Standard Deviation ($\sigma$), Mean, Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), however as remission is a ratio of minimum to maximum detectable reflected intensity there was no need to render the features non-dimensionalized. Thirty percent of the training data was randomly selected and reserved to create a validation data set.}

@@

			{MATLAB's Classification Learner Application was used to generate a Random Decision Forests for each considered scanning LiDAR channel. Decision tree depth, number of learners, and number of features to consider for each binary split are hyper-parameters that were optimized using MATLAB's Bayesian Optimization. Bayesian Optimization automates the manual hyper-parameter tuning process by creating a surrogate model representing the objective function, in this case out-of-bag error. Initialization of the objective function is completed by training a number of RDFs using randomly selected hyper-parameters. Out-of-bag error is then found and used to create a series of Gaussian curves. Additional hyper-parameters are chosen by an acquisition function to determine their utility in minimizing out-of-bag error. Best hyper-parameters are chosen after a set number of iterations and used to create a final RDF. Confusion matrixes were generating showing the out-of-bag error to determine the training error (Fig. \ref{fig:out_of_bag_err_conf_mat}). Validation of the RDF was completed using the reserved thirty percent of the training data (Fig. \ref{fig:vali_err_conf_mat}). This process was completed for each of the three considered Scanning LiDAR channels used for classification.}

			\begin{figure}[H]
				\centering
				\includegraphics[width=0.75\linewidth]{figures/c2_min_class_error}
				\caption[RDF Training Classification Error]{Classification error during Bayesian Optimization tuning.}
				\label{fig:c2_min_class_error}
			\end{figure}

			\begin{figure}[H]
				\centering
				\includegraphics[width=0.75\linewidth]{figures/out_of_bag_err_conf_mat}
				\caption[Out-of-Bag Error]{During training the RDF may be tested using the out-of-bag error. Confusion matrices are used to indicate the final RDF iteration training error.}
				\label{fig:out_of_bag_err_conf_mat}
			\end{figure}

			\begin{figure}[H]
				\centering
				\includegraphics[width=0.75\linewidth]{figures/vali_err_conf_mat}
				\caption[Validation Error]{Confusion matrix with the validation training data set.}
				\label{fig:vali_err_conf_mat}
			\end{figure}

@@

			\begin{figure}[H]
				\centering
				\includegraphics[width=0.75\linewidth]{figures/raw_classification_results}
				\caption[Classified Point Cloud]{Classified Point Cloud.}
				\label{fig:raw_classification_results}
			\end{figure}

			\begin{figure}[H]
				\centering
				\includegraphics[width=0.9\linewidth]{figures/rm_db_4_ToC_2}
				\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed and compared to actual road-surface areas. }
				\label{fig:rm_db_4_toc}
			\end{figure}

@@

		\begin{figure}[H]
			\centering
			\includegraphics[width=0.75\linewidth]{figures/dirt_v_gravel}
			\caption[Re-grassed Dirt vs Gravel]{Side-of-road detection accuracy was diminished due to overlapping features between dirt and gravel. }
			\label{fig:dirt_v_gravel}
		\end{figure}

@@

		% Summary of Work & Closing Statements
		{High definition scanning LiDAR and GPS data was gathered on Blackburn Road and an intercepting gravel lot in Athens Ohio. Properties describing surface roughness and remission data were extracted and used to train Random Decision Forests. Terrain classification of consecutive scans compared to manually defined road surface areas yielded accuracy scores for gauging algorithm capability for gravel surface detection. It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively at an estimated rate of 13.19 ms per 360 degree scan. Classified point clouds were manually examined and gravel areas were projected and compared with actual gravel surface areas. Overlapping results between manually projected and actual road surface areas resulted in 91.67\% intercepting gravel road detection accuracy. In conjunction with the overlapping areas and the high true-positive detection accuracy of gravel surfaces, this method of gravel road detection is a promising method of gravel road surface detection. This work addressed the problem of road surface detection on unmarked gravel and chipseal roads. Current LiDAR and Camera based detection models were insufficient due to lack of distinct features typical of urban roads, such as painted line markings or curbs, or have excessive computational or storage requirements. This work balanced accuracy and efficiency by using less intensive analysis techniques of smaller point cloud data sets. The described work was a method of detecting physical unmarked gravel and chipseal roads by using a terrain classification approach to predicting road surface area. The impact of this work is that autonomous vehicles using LiDAR may be able to detect unmarked gravel and chipseal road surfaces, allowing autonomous operations on 1.5 million miles of previously undetected rural roads.}

>>>>

	{Gravel roads comprise 34.8\% \cite{road_stats_2} of all road surfaces in the United States. Some nations have predominately unpaved road networks, such as India, where 70.7\% of all roads by mile are categorized as unpaved \cite{malik_lal_2019}. Detection of gravel roads may allow for expansion of current civilian autonomous vehicle capabilities as well as introduce opportunities for industries that operate in rural areas. Global Positioning Service (GPS) and high resolution maps should not be relied upon for navigation of rural roads alone as GPS is not accurate enough or may not have a reliable signal for precision driving \cite{noauthor_gpsgov_nodate}, and gravel roads are not marked or mapped. Systems that rely on cameras may fail when analyzing gravel roads, as these lack visual cues such as painted lane markings \cite{crisman_scarf_1993} and may in some circumstances closely resemble surrounding terrain. LiDAR point cloud processing systems depend upon distinct geometric features, most commonly curbs \cite{yadav_extraction_2017,liu_new_2013,qiu_fast_2016,fernandes_road_2014,seker_experiments_nodate,yang_semi-automated_2013,miyazaki_line-based_2014,hervieu_road_2013,smadja_road_nodate}, however as curbs are not installed on rural gravel roads \cite{skorseth_gravel_nodate} this method breaks down. Normal Distribution Transform (NDT) scan matching analytically compares two point cloud data sets to track current position, however this model relies on distinct geometric features, and suffers when distinguishing between terrain types making it insufficient for determining road boundaries \cite{biber_normal_2003}. Using 2D projection of planes unto a 3D point cloud of a road surface is an alternative approach, but likewise requires curbs or painted lines for road boundary definition \cite{fernandes_road_2014, borkar_robust_2009-1, guo_lane_2015}. Instead, using LiDAR may overcome difficulties in achieving detection of gravel roads by using terrain classification.}

	{Detection of unmarked road surfaces may utilize roughness properties without distinct markings or topographical features. Surface roughness is a measurable property that may be used to characterize a gravel road, as they are typically consistent with gravel type used \cite{skorseth_gravel_nodate} which may be exploited in a search and compare function, and is distinct from chipseal, grass, and other ground types \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain}. Surface roughness properties may be derived from processing LiDAR spatial and remission data from an aerial and surface level perspective \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain, pollyea_experimental_2012,rychkov_computational_2012,lague_accurate_2013,brubaker_use_2013,turner_estimation_2014,campbell_lidar-based_2017,shepard_roughness_2001,tegowski_statistical_2016,sock_probabilistic_2016,milenkovic_roughness_2018,yadav_extraction_2017, yadav_rural_2018}. Implementing a Random Decision Forests (RDF) terrain classification model using scanning LiDAR data to exploit surface roughness was studied for road surface detection in this study. Evaluation of the detection model will rely principally upon accuracy and processing time, as a moving vehicle must obtain reliable information of the road surface in a timely manner in order to inform trajectory updates, or else risk suffering an accident.}

@@

	{Current LiDAR based road surface detection models may be categorized into two methodologies. First is the analysis of road surface properties, which considers aspects such as topology and surface roughness. Detection based on road surface smoothness was studied by segmenting the point cloud into candidate road surface regions and searching for elevation jumps \cite{liu_new_2013}. Secondly is the detection of roadside curbs, which relies upon the height difference between the road surface and the curb for road edge detection. LiDAR point density at a curb's upper and lower edges can be used to indicate road edges \cite{ibrahim_curb-based_2012}. Fernandes et al. propose a road surface detection method using LiDAR \cite{fernandes_road_2014}, by projecting a 2D reference plane unto the 3D LiDAR data. In this study, road surfaces are assumed to be flat regions between two elevated regions such as curbs. Cameras have been used to supplement LiDAR point cloud data, which can implement 3D colored elevation maps to match pre-stored geometry data in a modified Iterative Closest Point (ICP) algorithm \cite{manz_detection_2011}. Supervised Classification Applied to Road Following (SCARF) is an algorithm that detects road surfaces based on color differences between the road and the surrounding terrain \cite{crisman_scarf_1993}. It was found that difficulties arise when detecting road surfaces in less colored environments \cite{crisman_scarf_1993,manz_detection_2011}. Edmond et al. propose a method to detect terrain type through implementation of vehicle vibration, however this method is limited to what the vehicle is currently driving over and thus cannot detect intercepting gravel roads \cite{dupont_online_2008}.}

@@

			{Detection of unmarked gravel road surfaces using terrain classification requires building a data base of terrain features, training an RDF, and testing the algorithm for accuracy of unmarked road detection (outline of work in Fig. \ref{fig:flowz_2}). Training data was gathered by the Ohio University Autonomous Van van (Fig. \ref{fig:Experimental_Apperatus}). MATLAB was used to extract and parse LiDAR and GPS data. Scanning LiDAR data was aggregated into a large point cloud by using GNSS and IMU data to determine each LiDAR scan's position and orientation in a dead-reckoning process. Aggregated point cloud data was manually examined and compared to satellite and camera images to determine and project road boundaries. Training data was extracted from a single arc per 360 degree scanning LiDAR channel if the arc was coincident to a manually projected 2D training data area. Random Decision Forests were trained and verified using MATLAB's Classifier App. Scanning LiDAR data was consecutively classified and compared to manually defined truth areas yielded the true-positive accuracy for gravel, asphalt, and non-road surfaces. Classified point cloud data was examined without reference to truth areas in order to manually project possible road surfaces. Manually projected road surface areas were compared to truth areas to determine the accuracy of asphalt road and intercepting gravel driveway detection.}

@@

			{Required physical data was gathered by the Ohio University Autonomous Van (Fig. \ref{fig:Experimental_Apperatus}). LiDAR point cloud data was gathered with the vehicle's roof-mounted Velodyne VLP-32C. Velodnyne's VLP-32C has 32 channels producing 300,000 points per second with a vertical field of view from -45$^{\circ}$ to $+$15$^{\circ}$, providing a minimum 0.33$^{\circ}$ vertical angular resolution \cite{vlp_32c} and 0.2$^{\circ}$ horizontal angular resolution. GPS data was recorded with a Novatel PwrPak 7D-E2, a Global Navigation Satellite System (GNSS) \& Inertial Navigation System (INS), and two GNSS-502 antennas by NavtechGPS. Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using the Robot Operating System (ROS). Scanning LiDAR data of a gravel surface was gathered from a gravel parking lot that intersects Blackburn Road near Athens Ohio. Scanning LiDAR data of an asphalt surface was gathered from Blackburn Road near the intersecting gravel lot. Scanning LiDAR data of grassy surfaces was gathered from the side of the gravel lot. Although this work does not evaluate the detection accuracy of grass, differentiation between a road and non-road surface is necessary for evaluating road-surface detection, therefore grass is labeled as "unknown" in this work. Raw data was gathered using the Robotic Operating System (ROS) to package incoming scanning LiDAR, GNSS, IMU, and camera sensor data as rosbags. Rosbags were unpacked using MATLAB tools to separate data streams. Point cloud, GPS, and IMU topics were defined, and point cloud Cartesian coordinates with corresponding intensities, longitude, latitude, altitude, roll, pitch, and yaw messages were read from the appropriate topics and imported}

@@

			{Scanning LiDAR training data requires extraction from manually defined gravel, asphalt, and "unknown" terrain areas on aggregated point clouds. Aggregating LiDAR data into a single point cloud was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road. Scanning LiDAR timestamps were matched to the closest GNSS and IMU timestamp data. Closest matching GNSS and IMU data informed the location and orientation of the LiDAR scan. Point cloud translation and rotation was accomplished using transformation matrices derived from GPS and IMU data. Rotation matrices were created using extracted roll, pitch, and yaw data from the IMU. LiDAR origin was found using GPS longitude, latitude, and altitude data. Physical distances between the GPS, IMU, and LiDAR sensors was rectified by obtaining reference frames provided by AutonomouStuff. GPS coordinates were offset by the current orientation and converted the ground truth to the LiDAR frame. GPS, IMU, and LiDAR reference frames and rotational updates were combined into trajectory vectors. Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Fig. \ref{fig:road_areas_annotated_22}). While not as robust as more sophisticated methods of point cloud aggregation such as NDT or ICP scan matching, over shorter distances using this method proved to be adequate for this work.}

			\begin{figure}[H]
				\centering
				\includegraphics[width=0.75\linewidth]{figures/blackburn_road}
				\caption[Blackburn Road Camera View]{Camera view of Blackburn Road - an unmarked asphalt road with three intercepting gravel driveways from which asphalt data was gathered.}
				\label{fig:Blackburn_Road_View}
			\end{figure}

			\begin{figure}[H]
				\centering
				\includegraphics[width=0.75\linewidth]{figures/gravel_lot_pic}
				\caption[Gravel Training Lot]{Camera view of the gravel lot from which gravel training data was gathered.}
				\label{fig:gravel_training_lot}
			\end{figure}

			\begin{figure}[H]
				\centering
				\includegraphics[width=0.75\linewidth]{figures/pc_example}
				\caption[Aggregated Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map.}
			\end{figure}

			{Manually defined areas may then be overlayed unto the point cloud (Fig. \ref{fig:road_areas_annotated_22}). Six degree arcs (Fig. \cite{fig:area_example}) from the first three channels of scanning LiDAR data that lay directly in front of the vehicle were extracted and exported to a database if coincident with the manually defined areas. Training data was extracted from a single arc from each 360 degree LiDAR scan if the arc was coincident to a manually projected 2D area. Raw spatial and intensity values from the coincident arc were saved to a raw training database. Separate training databases were maintained for each individual channel. Features that describe the raw spatial and intensity data were extracted. Thirty percent of the training data base was randomly selected and set aside as a validation data base. Training data was visually examined for clustering to verify that a Random Decision Forest would be able to be trained (Fig. \ref{fig:training_data_cluster_2}). It was found that each class generally clustered to a certain range with some overlap, which indicates that a Random Decision Forest may be trained.}

			\begin{figure}[H]
				\centering
				\includegraphics[width=0.75\linewidth]{figures/road_areas_annotated_22}
				\caption[Manually Classified Point Cloud Data]{ Projected manually defined areas define terrain surface type. Three gravel driveways (annotated 1-3) intercept an asphalt road, area $3$ is the gravel parking lot entrance from which training data was extracted.}
				\label{fig:road_areas_annotated_22}
			\end{figure}

			\begin{figure}[H]
			\centering
			\includegraphics[width=0.9\linewidth]{figures/training_data_cluster_3}
			\caption[Example Clustering]{Example of training data comparison between two features.}
			\label{fig:training_data_cluster_2}
			\end{figure}

@@

			{LiDAR returns spatial data represented by Cartesian coordinates with units in meters and remission data represented by a dimensionless ratio of minimum to maximum detected brightness. Feature extraction was completed by performing mathematical functions on the gathered spatial training data. Standard Deviation ($\sigma$), Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), where $G$ is an $1*n$ array comprising of the differences between consecutive points in the spatial data array. Spatial features were non-dimensionalized as necessary by dividing by the average range of the arc to LiDAR point of origin to the appropriate power. Remission features include Standard Deviation ($\sigma$), Mean, Roughness ($max - min$), Min-Max Ratio ($min / max$), Min$^{2}$-Max Ratio ($min^2 / max$), and Gradient ($sqrt(\sum_{1}^{n} G))$), however as remission is a ratio of minimum to maximum detectable reflected intensity there was no need to render the features non-dimensionalized. Thirty percent of the training data was randomly selected and reserved to create a validation data set.}

@@

			{MATLAB's Classification Learner Application was used to generate a Random Decision Forests for each considered scanning LiDAR channel. Decision tree depth, number of learners, and number of features to consider for each binary split are hyper-parameters that were optimized using MATLAB's Bayesian Optimization. Bayesian Optimization automates the manual hyper-parameter tuning process by creating a surrogate model representing the objective function, in this the upper confidence interval of the classification error objective model (Fig. \ref{fig:c2_min_class_error}). Initialization of the objective function is completed by training a number of RDFs using randomly selected hyper-parameters. Out-of-bag error is then found and used to create a series of Gaussian curves. Additional hyper-parameters are chosen by an acquisition function to determine their utility in minimizing out-of-bag error. After a set number of iterations, best hyper-parameters are chosen based minimum upper confidence interval of the classification error objective model. Hyperparameter tuning may lead to over-fitting to the training data base. Testing for over-fitting was accomplished by comparing the out-of-bag training error to the validation error (Fig. \ref{fig:train_vs_valid_overfit_test2}). Confusion matrices provide classification accuracy information on training and validation data sets by comparing actual and predicted classes (Fig. \ref{fig:out_of_bag_err_conf_mat}, \ref{fig:vali_err_conf_mat}). This process was completed for each of the three considered scanning LiDAR channels used for classification.}

			\begin{figure}[H]
				\centering
				\includegraphics[width=0.75\linewidth]{figures/c2_min_class_error}
				\caption[RDF Training Classification Error]{Classification error during Bayesian Optimization tuning. Best hyper-parameters are chosen based minimum upper confidence interval of the classification error objective model, not minimum classification error.}
				\label{fig:c2_min_class_error}
			\end{figure}

			\begin{figure}[H]
				\centering
				\includegraphics[width=0.75\linewidth]{figures/train_vs_valid_overfit_test2}
				\caption[Training vs Validation Error]{Hyperparameter tuning may lead to over-fitting. Testing out-of-bag versus validation error demonstrates that the model is not over-fitting to the training data base if the validation error does not diverge from the out-of-bag error.}
				\label{fig:train_vs_valid_overfit_test2}
			\end{figure}

			\begin{figure}[H]
				\centering
				\includegraphics[width=0.75\linewidth]{figures/chan_2c_conf_OOB_mat222}
				\caption[Out-of-Bag Error]{During training the RDF may be tested using the out-of-bag error. Confusion matrices are used to indicate the final RDF iteration training error.}
				\label{fig:out_of_bag_err_conf_mat}
			\end{figure}

			\begin{figure}[H]
				\centering
				\includegraphics[width=0.75\linewidth]{figures/chan_2c_conf_VALIDATION_mat222}
				\caption[Validation Error]{Confusion matrix with the validation training data set.}
				\label{fig:vali_err_conf_mat}
			\end{figure}

@@

			\begin{figure}[H]
				\centering
				\includegraphics[width=0.9\linewidth]{figures/raw_classification_results_v3}
				\caption[Classified Point Cloud]{Classified Point Cloud.}
				\label{fig:raw_classification_results}
			\end{figure}

			\begin{figure}[H]
				\centering
				\includegraphics[width=0.9\linewidth]{figures/rm_db_4_ToC_22}
				\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed and compared to actual road-surface areas. }
				\label{fig:rm_db_4_toc}
			\end{figure}

@@

		\begin{figure}[H]
			\centering
			\includegraphics[width=0.75\linewidth]{figures/dirt_v_gravel2}
			\caption[Re-grassed Dirt vs Gravel]{Side-of-road detection accuracy was diminished due to overlapping features between dirt and gravel. }
			\label{fig:dirt_v_gravel}
		\end{figure}

@@

		% Summary of Work & Closing Statements
		{High definition scanning LiDAR and GPS data was gathered on Blackburn Road and an intercepting gravel lot in Athens Ohio. Properties describing surface roughness and remission data were extracted and used to train Random Decision Forests. Terrain classification of consecutive scans compared to manually defined road surface areas yielded accuracy scores for gauging algorithm capability for gravel surface detection. It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively at an estimated rate of 13.19 milliseconds per 360 degree scan. Classified point clouds were examined and gravel areas were projected and compared with actual gravel surface areas. Overlapping results between detected and actual road surface areas resulted in 91.67\% intercepting gravel road detection accuracy. In conjunction with the overlapping areas and the high true-positive detection accuracy of gravel surfaces, this method of gravel road detection is promising. This work addressed the problem of road surface detection on unmarked gravel and chipseal roads. The described work was a method of detecting physical unmarked gravel and chipseal roads by using a terrain classification approach to predicting road surface area. The impact of this work is that autonomous vehicles using LiDAR may be able to detect unmarked gravel and chipseal road surfaces, allowing autonomous operations on 1.5 million miles of previously undetected rural roads.}
